Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups

### 2022-01-14 manuscript/manuscript.Rmd
results

output: rticles::elsevier_article
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("output"),
      output_file = glue::glue("{xfun::sans_ext(input)}_{Sys.Date()}.pdf"),
      envir = globalenv()
    )
  })
---

@@

The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health. They include the amount of daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022. For each day during this period, school closure either due to a government decision or a public holiday is determined.

@@

@@

By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAMs allow the identification of dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:

@@

Contrary to Granger Causality, Transfer Entropy is robust to non-stationary time series [@silini2021fast]. The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups is done using the R package RTransferEntropy [@behrendt2019rtransferentropy].

@@

>>>>

header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}

output:
  bookdown::pdf_book:
    base_format: rticles::elsevier_article
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = here::here("output"),
      output_file = glue::glue("{xfun::sans_ext(input)}_{Sys.Date()}.pdf"),
      envir = globalenv()
    )
  })
---

(ref:overall) Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.
(ref:descriptive) Day-by-day changes in Covid-19 cases by age groups in Ireland: raw data (bars) and 7 aay rolling average (lines).

```{r setup, include = FALSE}
# libraries --------------------------------------------------------------------
library(here)
library(janitor)
library(knitr)
library(tidyverse)
library(zoo)

# options ----------------------------------------------------------------------
set.seed(123) # Seed for random number generation
options(scipen = 999) # disable sci number format
opts_chunk$set(
  cache.extra = rand_seed, 
  message = FALSE, 
  warning = FALSE, 
  error = FALSE,
  echo = FALSE, 
  fig.retina = 3
  )

# variables --------------------------------------------------------------------
age_order <- c("aged1to4", "aged5to14", "aged15to24", "aged25to34", "aged35to44", "aged45to54", "aged55to64", "aged65to74", "aged75to84", "aged85up")
age_name  <- c("1 to 4",   "5 to 14",   "15 to 24",   "25 to 34",   "35 to 44",   "45 to 54",   "55 to 64",   "65 to 74",   "75 to 84",   "85 and up")
age_color <- c("#999999",  "#E69F00",   "#56B4E9",    "#009E73",    "#F0E442",    "#0072B2",    "#D55E00",    "#CC79A7",    "#000000",    "#CAB2D6")

# data -------------------------------------------------------------------------
df <- 
  read_csv(
    here("data/age_groups.csv"), 
    col_types = list(
      .default = col_double(), 
      Date = col_date(format = "%d/%m/%Y")
    )
  ) |> 
  clean_names() |> 
  rename(total_cases = total) |> 
  arrange(date) |> 
  mutate(
    timeline = date - lag(date),  # check if gap between dates
    total_changes = total_cases - lag(total_cases)
  )

df_long <- df |> 
  select(c(date, starts_with("aged"))) |> 
  pivot_longer(-date, names_to = "age_group", values_to = "covid_cases") |> 
  mutate(age_group = factor(age_group, levels = age_order)) |> 
  group_by(age_group) |> 
  mutate(
    covid_cases_7davg = rollmean(covid_cases, k = 7, fill = NA),
    covid_changes = covid_cases - lag(covid_cases),
    covid_changes_7davg = rollmean(covid_changes, k = 7, fill = NA)
  ) |> 
  ungroup()
```

@@

The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health. They include the amount of daily cases for 10 age groups from `r format(min(df$date), "%B %d, %Y")` to `r format(max(df$date), "%B %d, %Y")`. For each day during this period, school closure either due to a government decision or a public holiday is determined.

@@

### Generalised Additive Model

@@

By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation [@wood2011fast], GAM identifies dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:

@@

### Transfer Entropy

Transfer Entropy can be used to infer the temporal relationship between two time series $I$ and $J$. This measure indicates whether $J$ can be used to to reduce the uncertainty on the future of $I$ and, consequently, that $J$ causes $I$ [@schreiber2000measuring]. Contrary to Wiener-Granger Causality, Transfer Entropy is robust to non-stationary time series [@wollstadt2014efficient; @papana2016detecting]. The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups using Shannon's calculation [@shannon1948mathematical] is done using the R package RTransferEntropy [@behrendt2019rtransferentropy].

The influence of day-by-day changes in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:

\begin{align}
  T_{J \rightarrow I}(k,l) = \sum_{i,j} p\left(i_{t+1}, i_t^{(k)}, j_t^{(l)}\right) \cdot log \left(\frac{p\left(i_{t+1}| i_t^{(k)}, j_t^{(l)}\right)}{p\left(i_{t+1}|i_t^{(k)}\right)}\right),
\end{align}

\noindent where $T_{J\rightarrow I}$ consequently measures the influence of the change dynamic from an eage group $J$ to another age group $I$.

@@

Overall, the trend of comfirm Covid-19 cases in Ireland is similar to those of other European countries (Figure \@ref(fig:overall)). Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).

```{r overall, fig.width=4.5, fig.height=3, fig.cap="(ref:overall)", fig.pos = "!h"}
overall_plot <- df |> 
  ggplot(aes(date, total_cases)) +
  geom_tile(aes(fill = factor(schools_closed), height = Inf), na.rm = TRUE, alpha = .4) +
  geom_line(na.rm = TRUE) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b %y") +
  scale_fill_manual(labels = c("Open", "Closed", "Na"), values = c("light blue", "red", "orange")) +
  labs(
    x = "Date", 
    y = "Covid-19 Cases",
    fill = "Schools"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 10, family = "serif"),
    axis.text = element_text(size = 10),
    panel.background = element_rect(colour = "black", size = 0.1)
  )

overall_plot
```

The evolution of day-by-day changes in Covid-19 cases reveals some similarities across all age groups. However, the influence of each waves on the each groups has also some particularities (Figure \@ref(fig:descriptive)). For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.

```{r descriptive, fig.width=4.5, fig.height=7, fig.cap="(ref:descriptive)"}
descriptive_plot <- df_long |> 
  ggplot() +
  geom_col(aes(date, covid_changes), na.rm = TRUE) +
  geom_line(aes(date, covid_changes_7davg, color = age_group), na.rm = TRUE) +
  scale_x_date("Date", date_breaks = "3 months", date_labels = "%b %y") +
  scale_y_continuous("Day-by-day Changes in Covid-19 Cases") +
  scale_color_manual(labels = age_name, values = age_color) +
  facet_grid(
    age_group ~ ., 
    scales = "free",
    labeller = as_labeller(
      c(
        aged1to4 = "1 to 4",
        aged5to14 = "5 to 14",
        aged15to24 = "15 to 24", 
        aged25to34 = "25 to 34", 
        aged35to44 = "35 to 44", 
        aged45to54 = "45 to 54", 
        aged55to64 = "55 to 64", 
        aged65to74 = "65 to 74", 
        aged75to84 = "75 to 84", 
        aged85up = "85 and up"
        )
      )
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.background = element_blank(),
    strip.placement = "outside",
    text = element_text(size = 10, family = "serif"),
    axis.text = element_text(size = 10),
    panel.background = element_rect(colour = "black", size = 0.1)
  )

descriptive_plot
```

### 2022-01-14 output/manuscript_2022-01-14.tex
results

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

@@

% Pandoc header

@@

The data were collected from the official daily publication of Covid-19
cases by the Irish Department of Health. They include the amount of
daily cases for 10 age groups from March 4th, 2020 to January 1st, 2022.
For each day during this period, school closure either due to a
government decision or a public holiday is determined.

@@

Four periods of school closure longer that 14 days have been identified.
These four periods are used to fit a Generalised Additive Model or GAM
(Wood, 2017) in order to test the hypothesis of a significant change in
the evolution of cases among age groups from 1 to 4, from 5 to 14, and
from 15 to 24.

By estimating the degree of smoothness of a Bayesian spline smoothing
using restricted maximum likelihood estimation (Wood, 2011), GAMs allow
the identification of dynamic patterns underlying the day-by-day changes
in Covid-19 cases reported while taking into account the random effect
of different age groups as follows:

@@

Contrary to Granger Causality, Transfer Entropy is robust to
non-stationary time series (Silini and Masoller, 2021). The Transfer
Entropy analysis of younger age groups daily changes in Covid-19 cases
on older age groups is done using the R package RTransferEntropy
(Behrendt et al., 2019).

@@

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-behrendt2019rtransferentropy}{}%
Behrendt, S., Dimpfl, T., Peter, F.J., Zimmermann, D.J., 2019.
RTransferEntropy --- quantifying information flow between different time
series using effective transfer entropy. SoftwareX 10, 100265.
doi:\href{https://doi.org/10.1038/s41598-021-87818-3}{10.1038/s41598-021-87818-3}

@@

@@

\leavevmode\hypertarget{ref-silini2021fast}{}%
Silini, R., Masoller, C., 2021. Fast and effective pseudo transfer
entropy for bivariate data-driven causal inference. Scientific Reports
11, 1--13.
doi:\href{https://doi.org/10.1016/j.softx.2019.100265}{10.1016/j.softx.2019.100265}

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Influence of School Closure on COVID-19 Contaminations in Ireland and Repercussions across Age Groups},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

@@

% Pandoc header
\usepackage{booktabs}
\usepackage{float}

@@

The data were collected from the official daily publication of Covid-19 cases by the Irish Department of Health. They include the amount of daily cases for 10 age groups from March 04, 2020 to January 01, 2022. For each day during this period, school closure either due to a government decision or a public holiday is determined.

@@

\hypertarget{generalised-additive-model}{%
\subsubsection{Generalised Additive Model}\label{generalised-additive-model}}

Four periods of school closure longer that 14 days have been identified. These four periods are used to fit a Generalised Additive Model (or GAM) using the R package mgcv (Wood, 2017) in order to test the hypothesis of a significant change in the evolution of cases among age groups from 1 to 4, from 5 to 14, and from 15 to 24.

By estimating the degree of smoothness of a Bayesian spline smoothing using restricted maximum likelihood estimation (Wood, 2011), GAM identifies dynamic patterns underlying the day-by-day changes in Covid-19 cases reported while taking into account the random effect of different age groups as follows:

@@

\hypertarget{transfer-entropy}{%
\subsubsection{Transfer Entropy}\label{transfer-entropy}}

Transfer Entropy can be used to infer the temporal relationship between two time series \(I\) and \(J\). This measure indicates whether \(J\) can be used to to reduce the uncertainty on the future of \(I\) and, consequently, that \(J\) causes \(I\) (Schreiber, 2000). Contrary to Wiener-Granger Causality, Transfer Entropy is robust to non-stationary time series (Papana et al., 2016; Wollstadt, 2014). The Transfer Entropy analysis of younger age groups daily changes in Covid-19 cases on older age groups using Shannon's calculation (Shannon, 1948) is done using the R package RTransferEntropy (Behrendt et al., 2019).

The influence of day-by-day changes in Covid-19 cases across all age groups is evaluated using Shannon's transfer entropy is given by:

\begin{align}
  T_{J \rightarrow I}(k,l) = \sum_{i,j} p\left(i_{t+1}, i_t^{(k)}, j_t^{(l)}\right) \cdot log \left(\frac{p\left(i_{t+1}| i_t^{(k)}, j_t^{(l)}\right)}{p\left(i_{t+1}|i_t^{(k)}\right)}\right),
\end{align}

\noindent where \(T_{J\rightarrow I}\) consequently measures the influence of the change dynamic from an eage group \(J\) to another age group \(I\).

@@

Overall, the trend of comfirm Covid-19 cases in Ireland is similar to those of other European countries (Figure \ref{fig:overall}). Indeed it is possible to observe a significant increase corresponding to each of the 4 waves of transmission (March-April 2020, December 2020-January 2021, September-October 2021 and November-December 2021).

\begin{figure}[!h]
\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_2022-01-14_files/figure-latex/overall-1} \caption{Evolution of confirmed Covid-19 cases in Ireland alongside with school closures.}\label{fig:overall}
\end{figure}

The evolution of day-by-day changes in Covid-19 cases reveals some similarities across all age groups. However, the influence of each waves on the each groups has also some particularities (Figure \ref{fig:descriptive}). For example, the first wave was more important among the oldest age groups whereas the third wave was more important among the youngest age groups.

\begin{figure}
\includegraphics[width=324px]{/Users/damiendcu/OneDrive/Projects/covid_ts_causality/output/manuscript_2022-01-14_files/figure-latex/descriptive-1} \caption{Day-by-day changes in Covid-19 cases by age groups in Ireland: raw data (bars) and 7 aay rolling average (lines).}\label{fig:descriptive}
\end{figure}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-behrendt2019rtransferentropy}{}%
Behrendt, S., Dimpfl, T., Peter, F.J., Zimmermann, D.J., 2019. RTransferEntropy --- quantifying information flow between different time series using effective transfer entropy. SoftwareX 10, 100265. doi:\href{https://doi.org/10.1016/j.softx.2019.100265}{10.1016/j.softx.2019.100265}

@@

\leavevmode\hypertarget{ref-papana2016detecting}{}%
Papana, A., Kyrtsou, C., Kugiumtzis, D., Diks, C., 2016. Detecting causality in non-stationary time series using partial symbolic transfer entropy: Evidence in financial data. Computational economics 47, 341--365. doi:\href{https://doi.org/10.1007/s10614-015-9491-x}{10.1007/s10614-015-9491-x}

\leavevmode\hypertarget{ref-schreiber2000measuring}{}%
Schreiber, T., 2000. Measuring information transfer. Phys. Rev. Lett. 85, 461--464. doi:\href{https://doi.org/10.1103/PhysRevLett.85.461}{10.1103/PhysRevLett.85.461}

\leavevmode\hypertarget{ref-shannon1948mathematical}{}%
Shannon, C.E., 1948. A mathematical theory of communication. The Bell system technical journal 27, 379--423.

@@

@@

\leavevmode\hypertarget{ref-wollstadt2014efficient}{}%
Wollstadt, M.A.V., Patricia AND Martínez-Zarzuela, 2014. Efficient transfer entropy analysis of non-stationary neural time series. PLOS ONE 9, 1--21. doi:\href{https://doi.org/10.1371/journal.pone.0102833}{10.1371/journal.pone.0102833}
